{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-10 thesis/tm.tex
summarize tm chapater

\revision{If the user supplies the {\tt -W} option (\cref{sec:tm-interface}),
Landslide will emulate STM's weak atomicity semantics (\cref{sec:tm-weak-atomicity}).
%to allow non-transactional code
%to interleave between a {\tt \_xbegin()}/{\tt \_xend()} pair.
Concretely, it achieves this by
reporting both halves of data races that occur between transactional and non-transactional code (\cref{sec:landslide-datarace})
(previously, only the latter half would be reported),
allowing preemption points to be identified during transactions (\cref{sec:landslide-pps}),
and allowing system calls during transactions to proceed rather than forcing a {\tt CAPACITY} abort
(as programmers may now need to nest other synchronization within transactions to protect against non-transactional code).
%and traversing the interleaving arbitrarily far back from a {\tt \_xend()} to find the appropriate preemption
%(previously, the xbegin would be at the immediately previous PP)
Landslide requires the other STM options ({\tt -A -S}) also to be enabled to emulate weak atomicity,
because I know of no existing HTM system which implements it.
As one might expect, it inflates state spaces considerably, % TODO: measure it with htm2 or some such?
and moreover,
every single HTM benchmark in the upcoming evaluation,
implemented to rely on strong atomicity,
exhibits bugs under weak atomicity.
Commit 68401fc implements this mode.}

@@

Finally, I identified a specific pattern of transactional code where existing state-space reduction algorithms
will fail to identify and prune equivalent thread interleavings.
\Cref{fig:retry-set-example} shows a minimal example program which exhibits this problem.
In this program, each thread's transactional path conflicts with the other thread's abort path,
while the two transactional paths (disjoint memory accesses) and the two abort paths (reads only) are pairwise independent.

@@

% TODO: is this true? at all? htm2 is bigger in STM only because it's got a separate expliit abort
Likewise, the cases of {\tt htm2}(2,3), (2,4), and (4,1),
in which retry sets achieved better reduction than STM mode,
show that the latter does not necessarily subsume the former,
and that combining the two could in theory achieve further reduction still.
However, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
may depend on the execution logic
(explicit aborts may be conditional on some change by a conflicting thread,
and conflict aborts cannot occur before their conflicting transaction to begin with%
\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
already halfway into implementing retry sets the wrong way at first.}%
),
so how to soundly prune either success paths or retry aborts
while other abort codes are in play
remains an open problem.

@@

>>>>

\revision{If the user supplies the {\tt -W} option (\cref{sec:tm-interface}),
Landslide will emulate STM's weak atomicity semantics (\cref{sec:tm-weak-atomicity}).
%to allow non-transactional code
%to interleave between a {\tt \_xbegin()}/{\tt \_xend()} pair.
Concretely, it achieves this by
reporting both halves of data races that occur between transactional and non-transactional code (\cref{sec:landslide-datarace})
(previously, only the latter half would be reported),
allowing preemption points to be identified during transactions (\cref{sec:landslide-pps}),
and allowing system calls during transactions to proceed rather than forcing a {\tt CAPACITY} abort
(as programmers may now need to nest other synchronization within transactions to protect against non-transactional code).
%and traversing the interleaving arbitrarily far back from a {\tt \_xend()} to find the appropriate preemption
%(previously, the xbegin would be at the immediately previous PP)
Landslide requires the other STM options ({\tt -A -S}) also to be enabled to emulate weak atomicity,
partly because I know of no existing HTM system which implements it,
but mostly to keep the implementation simple.
As one might expect, it inflates state spaces considerably, % TODO: measure it with htm2 or some such?
and moreover,
every single HTM benchmark in the upcoming evaluation,
implemented to rely on strong atomicity,
exhibits bugs under weak atomicity.
Commit 68401fc implements this mode.}

@@

Finally, I identified a specific pattern of transactional code where existing state-space reduction algorithms
will fail to identify and prune equivalent thread interleavings.
\Cref{fig:retry-set-example} shows a minimal example program which exhibits this problem.
In this program, each thread's transactional path conflicts with the other thread's abort path,
while the two transactional paths (disjoint memory accesses) and the two abort paths (reads only) are
independent with each other.

@@

Likewise, the cases of {\tt htm2}(2,3), (2,4), and (4,1),
in which retry sets achieved better reduction than STM mode,
\revisionminor{suggest} that the latter does not necessarily subsume the former,
and that combining the two could in theory achieve further reduction still,
However, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
may depend on the execution logic
(explicit aborts may be conditional on some change by a conflicting thread,
and conflict aborts cannot occur before their conflicting transaction to begin with%
\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
already halfway into implementing retry sets the wrong way at first.}%
),
so \revisionminor{it remains an open problem}
to soundly prune either success paths or retry aborts
while other abort codes are in play,
\revisionminor{and preserve the reduction achieved with the baseline configuration}.

@@

\section{Summary}

\revision{This chapter has extended the concurrency model of stateless model checking to support transactional memory,
and proposed several new reduction strategies for coping with the concomitant % hi william
state space explosion.
I have made the following contributions:

\begin{itemize}
	\item Three equivalence proofs,
		demonstrating that model checkers can soundly emulate strongly-atomic HTM execution semantics by
		treating transactions as globally-uninterruptible atomic sections
		and injecting failure without simulating rollback
		(\cref{sec:tm-design-failure})
		and applying conventional data race analysis with all transactions treated as a global lock
		(\cref{sec:tm-design-locks}),
		and can emulate weakly-atomic STM systems by relaxing the above atomicity to apply to other transactions only
		(\cref{sec:tm-weak-atomicity}).
	\item An implementation of both forms of transactional memory in Landslide,
		with further options to configure the variety of abort codes injected
		(\cref{sec:tm-implementation}).
	\item A suite of transactional benchmark programs,
		designed to produce interesting state spaces of varying size, bugginess, and reducibility
		(\cref{sec:tm-eval-exp-setup}).
		This test suite is more appropriate for evaluating model checkers
		% get rekt STAMP :triumph:
		than prior work, which to date has focused on testing the raw performance of TM backend implementations,
		with little focus on the interactions of abort paths.
	\item Two previously-unknown bugs found in transactional programs from prior work
		(\cref{sec:tm-eval-bugs}).
	\item Retry set reduction,
		a new state space reduction algorithm
		(\cref{sec:tm-retrysets})
		which the evaluation shows identifies and skips equivalent interleavings
		compared to standard DPOR, even when transactional and abort paths are fully mutually conflicting
		(\cref{sec:tm-eval-retrysets}).
	\item Demonstration of two further heuristic reduction strategies on HTM programs,
		abort code reduction (\cref{sec:tm-eval-stm})
		and abstraction reduction (\cref{sec:tm-abstraction})
		which rely on human intuition to ensure their soundness
		to tackle even larger state spaces still.
	\item Verifications of \Cref{fig:htm-fixed}'s atomicity protocol %(with retry set reduction)
		and the {\tt lock}({\tt mutex}) transactional lock %(with abort code reduction)
		up to 2 threads, 4 iterations and 4 threads, 1 iteration for each
		(\cref{sec:tm-verif}).
		%$(K,N)$=$(2,4)$ and $(4,1)$ for each.
		% verified htm atomicity protocol (htm2) and mutex (lock mutex) up to (2,4), (4,1)
\end{itemize}

As concurrent systems grow in complexity to meet modern performance demands,
they grow ever more fraught with opportunity for subtle bugs,
and the programmer likewise faces ever more difficulty in verifying her programs.
Following in the footsteps of related work for relaxed memory concurrency,
this chapter has enabled stateless model checking to address yet another new technique,
and concludes this thesis's contributions to the field.
The upcoming chapters will offer a tour of related research problems,
solved and unsolved,
practical and philosophical,
past and future.
